PineAPPL:\\ EW and mixed QCD--EW corrections for PDF fits

### 2020-08-03 paper/sec-introduction.tex
Intro revised

With the recent completion of Run II, the Large Hadron Collider (LHC) has 
accumulated data from an integrated luminosity of approximately 
\SI{150}{\per\femto\barn}~\cite{Mangano:2020icy}. This represents only a small fraction of
the anticipated \SI{3000}{\per\femto\barn} that will eventually be recorded in the
forthcoming twenty years of LHC operation. Nevertheless the statistical 
uncertainty of the data has already shrunk to unprecedented small values,
typically \SI{1}{\percent} or less, a fact that will allow for precision tests of the
Standard Model (SM) and for indirect searches of New Physics only if 
theoretical predictions become comparatively precise. This entails the 
computation of additional higher-order contributions to the fixed-order 
coupling-constant perturbative expansion, on the one hand, and an increasingly 
sophisticated determination of the Parton Distribution Functions (PDFs) of the 
proton~\cite{Gao:2017yyd}, on the other hand.

In the first respect, because Quantum Chromodynamics (QCD) dominates the 
interactions occurring within colliding protons at the LHC, much effort 
has been devoted to the computation of higher-order QCD corrections: 
fully-differential next-to-leading order (NLO) results, possibly matched to the
parton shower, are currently automated in various general-purpose
Monte Carlo generators~\cite{Gleisberg:2008ta,Alwall:2014hca,Bellm:2015jjp}
(see also ref.~\cite{Buckley:2011ms} for a review),
while an increasing number of next-to-next-to-leading order (NNLO) predictions 
are becoming available for processes with various degrees of inclusiveness
(see e.g.\ ref.~\cite{Amoroso:2020lgh} and references therein). The computation
of higher-order corrections in the electroweak (EW) and mixed QCD+EW theory has
witnessed a comparatively less comprehensive progress. Nevertheless frameworks 
were developed in which the QCD and EW couplings are simultaneously treated as 
small parameters in the perturbative expansion, and the computation of 
theoretical predictions, accurate to NLO in both (including mixed-coupling 
QCD--EW terms), is automated~\cite{Biedermann:2017yoi,Frederix:2018nkq}.

In the second respect, contemporary PDF
sets~\cite{Harland-Lang:2014zoa,Ball:2017nwa,Hou:2019efy}
incorporate a significant amount of LHC data, which is analysed with NNLO QCD 
theory by default. No EW corrections are systematically included in the 
theoretical description of the experimental observables to which PDFs are 
optimised, except for QED effects if a photon PDF is 
determined~\cite{Schmidt:2015zda,Bertone:2017bme,Harland-Lang:2019pla}. 
The resulting relative PDF uncertainty---which accounts only for the
uncertainty of the data and of residual methodology inefficiencies inherent to 
each PDF determination---can be as low as 1\% at the EW
scale~\cite{Ball:2017nwa}. Theoretical uncertainties, possibly of comparable 
size (e.g.\ from missing higher-order terms in the QCD perturbative
expansion), have started to be included in a PDF determination and represented 
into PDF uncertainties only very 
recently~\cite{AbdulKhalek:2019bux,AbdulKhalek:2019ihb}. This has led to a 
noticeable improvement of PDF accuracy, in terms of an overall
better statistical quality of the fit, at the price of a modest deterioration 
of PDF precision, in terms of a slight inflation of PDF uncertainties.

@@

In this paper we address the first of these two problems. Specifically, 
we develop \textsc{PineAPPL}, a new library that provides a direct interface
from parton-level Monte Carlo event generators accurate up to NLO QCD+EW to
fast interpolation grids. The library supports variations of the factorisation
and renormalisation scales, resummation, and matching with a photon- and/or
parton-shower. The library is included by default in 
\textsc{MadGraph5\_aMC@NLO} (\textsc{mg5\_aMC@NLO} henceforth) within which it has
been designed and tested. However, it is suitable for use as a plugin in any 
Monte Carlo generator that supports NLO QCD+EW computations, such as 
\textsc{SHERPA}~\cite{Biedermann:2017yoi}. In this respect, \textsc{PineAPPL}
supersedes the \textsc{aMCfast} interface~\cite{Bertone:2014zva} and complements
the \textsc{MCgrid} interface~\cite{DelDebbio:2013kxa,Bothmann:2015dba}.

The paper is organised as follows. In section~\ref{sec:pineappl} we introduce
\textsc{PineAPPL}, we describe its features, we illustrate its operation, and we
assess its performance. In section~\ref{sec:results}, by means of
\textsc{mg5\_AMC@NLO} and \textsc{PineAPPL}, we compute fast interpolation grids for
a representative set of data that is typically included in a fit of PDFs.
In section~\ref{sec:doublecounting} we use these results to formulate more
comprehensively the second problem sketched above, the solution of which, 
however, remains beyond the scope of the current work. We provide our
conclusions and an outlook in section~\ref{sec:conclusion}.

%Outline a possible strategy to deal with EW correction in LHC observables 
%(FSR, PHOTOS, dressed/born, ...).

>>>>

With the recent completion of Run II, the Large Hadron Collider (LHC) has 
accumulated data from an integrated luminosity of approximately 
\SI{150}{\per\femto\barn}~\cite{Mangano:2020icy}. This represents only a small fraction of
the anticipated \SI{3000}{\per\femto\barn} that will eventually be recorded in the
forthcoming twenty years of LHC operation. Nevertheless the statistical 
uncertainty of the data has already shrunk to unprecedented small values,
typically \SI{1}{\percent} or less, a fact that will allow for precision tests of the
Standard Model (SM) and for indirect searches of New Physics only if 
theoretical predictions become comparatively precise. This entails the 
computation of additional higher-order contributions to the fixed-order 
coupling-constant perturbative expansion, on the one hand, and an increasingly 
sophisticated determination of the Parton Distribution Functions (PDFs) of the 
proton~\cite{Gao:2017yyd,Ethier:2020way}, on the other hand.

In the first respect, because Quantum Chromodynamics (QCD) dominates the 
interactions occurring within colliding protons at the LHC, much effort 
has been devoted to the computation of higher-order QCD corrections: 
fully-differential next-to-leading order (NLO) results, possibly matched to the
parton shower, are currently automated in various general-purpose
Monte Carlo generators~\cite{Gleisberg:2008ta,Alwall:2014hca,Bellm:2015jjp}
(see also ref.~\cite{Buckley:2011ms} for a review),
while an increasing number of next-to-next-to-leading order (NNLO) predictions 
are becoming available for processes with various degrees of inclusiveness
(see e.g.\ ref.~\cite{Amoroso:2020lgh} and references therein). The computation
of higher-order corrections in the electroweak (EW) and mixed QCD+EW theory has
also witnessed a comprehensive progress. Frameworks 
were developed in which the QCD and EW couplings are simultaneously treated as 
small parameters in the perturbative expansion, and the computation of 
theoretical predictions, accurate to NLO in both (including mixed-coupling 
QCD--EW terms), is automated~\cite{Biedermann:2017yoi,Frederix:2018nkq}.

In the second respect, contemporary PDF
sets~\cite{Harland-Lang:2014zoa,Ball:2017nwa,Hou:2019efy}
incorporate a significant amount of LHC data, which is analysed with NNLO QCD 
theory by default. No EW corrections are systematically included in the 
theoretical description of the experimental observables to which PDFs are 
optimised, except for QED effects if a photon PDF is 
determined~\cite{Schmidt:2015zda,Bertone:2017bme,Harland-Lang:2019pla}.
Data points in kinematic regions where EW corrections are expected to affect
the accuracy of the theoretical predictions are usually removed from the fits.
The resulting relative PDF uncertainty --- which accounts only for the
uncertainty of the data and of residual methodological inefficiencies inherent to 
each PDF determination --- can be as low as 1\% at the EW
scale~\cite{Ball:2017nwa}. Theoretical uncertainties, possibly of comparable 
size (e.g.\ from missing higher-order terms in the QCD perturbative
expansion), have started to be included in a PDF determination and represented 
into PDF uncertainties only very 
recently~\cite{AbdulKhalek:2019bux,AbdulKhalek:2019ihb}. This has led to an 
improvement of PDF accuracy, in terms of an overall
better statistical quality of the fit, at the price of a modest deterioration 
of PDF precision, in terms of a slight inflation of PDF uncertainties.

@@

In this paper we address the first of these two problems. Specifically, 
we develop \textsc{PineAPPL}, a new computer library that provides a direct interface
from parton-level Monte Carlo event generators accurate up to NLO QCD+EW to
fast interpolation grids. The library supports variations of the factorisation
and renormalisation scales, resummation, and matching with a photon- and/or
parton-shower. The library is included by default in 
\textsc{MadGraph5\_aMC@NLO} (\textsc{MG5\_aMC} henceforth) within which it has
been designed and tested. However, it is suitable for use as a plugin in any 
Monte Carlo generator that supports NLO QCD+EW computations, such as 
\textsc{SHERPA}~\cite{Biedermann:2017yoi}. In this respect, \textsc{PineAPPL}
supersedes \textsc{aMCfast}~\cite{Bertone:2014zva} and complements
\textsc{MCgrid}~\cite{DelDebbio:2013kxa,Bothmann:2015dba}.

The paper is organised as follows. In section~\ref{sec:pineappl} we introduce
\textsc{PineAPPL}, we describe its features, we illustrate its operation, and we
assess its performance. In section~\ref{sec:results} we validate
\textsc{PineAPPL} and demonstrate its capabilities by computing fast
interpolation grids, accurate to NLO QCD and NLO QCD+EW, for a representative
set of processes, usually included in PDF fits, for which EW corrections may
have a sizeable effect on the accuracy of the theoretical predictions.
In section~\ref{sec:doublecounting} we use these results to formulate more
comprehensively the second problem sketched above, the solution of which, 
however, remains beyond the scope of the current work. We provide our
conclusions and an outlook in section~\ref{sec:conclusion}. Examples of
installation and usage of \textsc{PineAPPL} are provided in
appendix~\ref{app:pineappl}; appendix~\ref{app:lumis} collects the parton
luminosities for each process considered in section~\ref{sec:results}; and
appendix~\ref{app:add_plots} complements some of the results presented
in section~\ref{sec:results}.
